---
title: "Homework Assignment #4"
subtitle: "Prioritizing Potential Aquaculture"
author: "Eric Fairbanks"
date: 11-20-2024
execute: 
  eval: true
format:
  html:
    toc: true
    warning: false
    error: false
    message: false
editor_options: 
  chunk_output_type: console
---

```{r}
library(tmap)
library(tidyverse)
library(dplyr)
library(sf)
library(here)
library(terra)
library(kableExtra)
```

### Part 1: Data Processing

### Load in Temp, Depth, and EEZs

```{r}
#load in data
sst_2008 <- rast(here("data", "average_annual_sst_2008.tif"))
sst_2009 <- rast(here("data", "average_annual_sst_2009.tif"))
sst_2010 <- rast(here("data", "average_annual_sst_2010.tif"))
sst_2011 <- rast(here("data", "average_annual_sst_2011.tif"))
sst_2012 <- rast(here("data", "average_annual_sst_2012.tif"))
depth <- rast(here("data", "depth.tif"))
eez <- st_read(here("data", "wc_regions_clean.shp")) #exclusive economic zones

#raster stack of sst
sst_stack <- c(sst_2008, sst_2009, sst_2010, sst_2011, sst_2012)
```

### Clean up Data

```{r}
#mean raster from sst raster_stack
mean_sst <- app(sst_stack, fun = mean) %>%
  project("EPSG:4326")
#convert to celsius
mean_sst <- mean_sst - 273.15

#crop depth to extens of sst
depth_new <- crop(depth, mean_sst) %>%
  resample(mean_sst, method = "near")

#check to see if depth and sst match
stack_check <- c(depth_new, mean_sst)
#success
```

### Part 2: Oysters

### Set Parameters and Reclassifications

```{r}
#set oyster paramaters
sst_min <- 11
sst_max <- 30
depth_min <- -70
depth_max <- -0


#resample sst to oyster conditions, 1 optimal, 0 not optimal
sst_rcl <- matrix(c(sst_min, sst_max, 1,
                       0, sst_min, 0,
                       sst_max, Inf, 0),
                     ncol = 3, byrow = TRUE)

x_sst <- classify(mean_sst, sst_rcl)

#resample depth to oyster conditions, 1 optimal, 0 not optimal
x_depth_rcl <- matrix(c(depth_min, depth_max, 1,
                             -Inf, depth_min, 0,
                             depth_max, Inf, 0),
                           ncol = 3, byrow = TRUE)

x_depth <- classify(depth_new, x_depth_rcl)
```

### Find Suitable Locations

```{r}
#use map algebra to find suitable locations
x_locations <- x_sst * x_depth
```

### Calculate Suitable Area for Each EEZ

```{r}
#rasterize eez data so we can do zonal analysis
eez_x_raster <- rasterize(eez, x_locations
                               , field = "rgn")

#calculate cell areas
cell_area <- cellSize(x_locations, unit = "m")

#combine suitable oyster locations with eez zones
x_eez_zones <- x_locations * eez_x_raster

#find total area of suitable space by eez
suitable_area_by_eez <- zonal(cell_area * (x_eez_zones > 0), eez_x_raster, fun = "sum", na.rm = TRUE)
```

### Visualize Findings

```{r}
#create dataframe that holds suitable area and geometry of eez
suitable_area_with_geometry <- suitable_area_by_eez %>%
  left_join(eez %>% select(rgn, geometry), by = "rgn") %>%
  mutate(area_km2 = area / 1e6) %>%
  st_as_sf()

#map of suitable area
tm_shape(suitable_area_with_geometry) +
  tm_fill(col = "area_km2", title = "Suitable Area (km²)", palette = "Blues") +
  tm_borders() +
  tm_layout(
    main.title = "Suitable Oyster Aquaculture Areas",
    title.position = c("center", "top"),
    main.title.size = 1,
    legend.position = c("left", "bottom"), 
    legend.outside = TRUE,  
    legend.outside.position = "right",
    legend.outside.size = 0.5
  ) +
  tm_scale_bar(position = c("left", "bottom")) +
  tm_compass(position = c("right", "top"),
             size = 1)

# create table of data
table_data <- suitable_area_with_geometry %>%
  st_drop_geometry() %>% 
  select(rgn, area_km2) %>%
  arrange(desc(area_km2)) 

table_data %>%
  kable(
    caption = "Suitable Oyster Habitat by Area (km²)",
    col.names = c("EEZ", "Suitable Area (km²)"),
    format = "html"
  ) %>%
   kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Part 3: Red Abalone

### Set Parameters and Reclassifications

```{r}
#set abalone paramaters
sst_min <- 8
sst_max <- 18
depth_min <- -24
depth_max <- -0


#resample sst to abalone conditions, 1 optimal, 0 not optimal
sst_rcl <- matrix(c(sst_min, sst_max, 1,
                       0, sst_min, 0,
                       sst_max, Inf, 0),
                     ncol = 3, byrow = TRUE)

x_sst <- classify(mean_sst, sst_rcl)

#resample depth to abalone conditions, 1 optimal, 0 not optimal
x_depth_rcl <- matrix(c(depth_min, depth_max, 1,
                             -Inf, depth_min, 0,
                             depth_max, Inf, 0),
                           ncol = 3, byrow = TRUE)

x_depth <- classify(depth_new, x_depth_rcl)
```

### Find Suitable Locations

```{r}
#use map algebra to find suitable locations
x_locations <- x_sst * x_depth
```

### Calculate Suitable Area for Each EEZ

```{r}
#rasterize eez data so we can do zonal analysis
eez_x_raster <- rasterize(eez, x_locations
                               , field = "rgn")

#calculate cell areas
cell_area <- cellSize(x_locations, unit = "m")

#combine suitable abalone locations with eez zones
x_eez_zones <- x_locations * eez_x_raster

#find total area of suitable space by eez
suitable_area_by_eez <- zonal(cell_area * (x_eez_zones > 0), eez_x_raster, fun = "sum", na.rm = TRUE)
```

### Visualize Findings

```{r}
#create dataframe that holds suitable area and geometry of eez
suitable_area_with_geometry <- suitable_area_by_eez %>%
  left_join(eez %>% select(rgn, geometry), by = "rgn") %>%
  mutate(area_km2 = area / 1e6) %>%
  st_as_sf()

#map of suitable area
tm_shape(suitable_area_with_geometry) +
  tm_fill(col = "area_km2", title = "Suitable Area (km²)", palette = "Blues") +
  tm_borders() +
  tm_layout(
    main.title = "Suitable Abalone Aquaculture Areas",
    title.position = c("center", "top"),
    main.title.size = 1,
    legend.position = c("left", "bottom"), 
    legend.outside = TRUE,  
    legend.outside.position = "right",
    legend.outside.size = 0.5
  ) +
  tm_scale_bar(position = c("left", "bottom")) +
  tm_compass(position = c("right", "top"),
             size = 1)

# create table of data
table_data <- suitable_area_with_geometry %>%
  st_drop_geometry() %>% 
  select(rgn, area_km2) %>%
  arrange(desc(area_km2)) 

table_data %>%
  kable(
    caption = "Suitable Abalone Habitat by Area (km²)",
    col.names = c("EEZ", "Suitable Area (km²)"),
    format = "html"
  ) %>%
   kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### Data Citations

\| Suitable growing conditions \| SeaLifeBase. 2024. Species Temperature and Depth Requirements for Marine Aquaculture. \| <https://www.sealifebase.ca/search.php> \|

\| Sea surface temperatures \| NOAA Coral Reef Watch. 2008-2012. 5km Daily Global Satellite Sea Surface Temperature Anomaly v3.1. \| <https://coralreefwatch.noaa.gov/product/5km/index_5km_ssta.php> \|

\| Bathymetry \| General Bathymetric Chart of the Oceans (GEBCO). 2024. Gridded Bathymetry Data. \| <https://www.gebco.net/data_and_products/gridded_bathymetry_data/#area> \|

\| Exclusive economic zones \| Marine Regions. 2024. Exclusive Economic Zones (EEZ) Boundaries, West Coast of the United States. \| <https://www.marineregions.org/eez.php> \|
