---
title: "Potential Marine Aquaculture"
author: "Eric Fairbanks"
date: 12-9-2024
execute: 
  eval: true
format:
  html:
    toc: true
    warning: false
    error: false
    message: false
editor_options: 
  chunk_output_type: console
---

```{r}
library(tmap)
library(tidyverse)
library(dplyr)
library(sf)
library(here)
library(terra)
library(kableExtra)
```

### Load in Temp, Depth, and EEZs

```{r}
#load in data
sst_2008 <- rast(here("data", "average_annual_sst_2008.tif"))
sst_2009 <- rast(here("data", "average_annual_sst_2009.tif"))
sst_2010 <- rast(here("data", "average_annual_sst_2010.tif"))
sst_2011 <- rast(here("data", "average_annual_sst_2011.tif"))
sst_2012 <- rast(here("data", "average_annual_sst_2012.tif"))
depth <- rast(here("data", "depth.tif"))
eez <- st_read(here("data", "wc_regions_clean.shp"), quiet = TRUE) #exclusive economic zones

#raster stack of sst
sst_stack <- c(sst_2008, sst_2009, sst_2010, sst_2011, sst_2012)
```

### Clean Up Data

```{r}
#mean raster from sst raster_stack
mean_sst <- app(sst_stack, fun = mean) %>%
  project("EPSG:4326")
#convert to celsius
mean_sst <- mean_sst - 273.15

#crop depth to extens of sst
depth_new <- crop(depth, mean_sst) %>%
  resample(mean_sst, method = "near")

#check to see if depth and sst match
stack_check <- c(depth_new, mean_sst)
#success
```

### Create Generalized Function

```{r}
suitable_locations <- function(mean_sst, depth_new, eez, 
                                  sst_min, sst_max, depth_min, depth_max, 
                                  title, palette = "Blues") {
  
  #reclassify sst
  sst_rcl <- matrix(c(sst_min, sst_max, 1,
                      0, sst_min, 0,
                      sst_max, Inf, 0), 
                    ncol = 3, byrow = TRUE)
  sst_suitable <- classify(mean_sst, sst_rcl)
  
  #reclassify depth
  depth_rcl <- matrix(c(depth_min, depth_max, 1,
                        -Inf, depth_min, 0,
                        depth_max, Inf, 0), 
                      ncol = 3, byrow = TRUE)
  depth_suitable <- classify(depth_new, depth_rcl)
  
  #identify suitable locations
  suitable_locations <- lapp(c(sst_suitable, depth_suitable), fun = function(x, y) x * y)
  
  #rasterize EEZ shapefile and calculate suitable area
  eez_raster <- rasterize(eez, suitable_locations, field = "rgn")
  cell_area <- cellSize(suitable_locations, unit = "m")
  eez_suitability <- suitable_locations * eez_raster
  suitable_area_by_eez <- zonal(cell_area * (eez_suitability > 0), eez_raster, fun = "sum", na.rm = TRUE)
  
  #merge results with EEZ shapefile
  suitable_area_sf <- suitable_area_by_eez %>%
    left_join(eez %>% select(rgn, geometry), by = "rgn") %>%
    mutate(area_km2 = area / 1e6)  %>%
    st_as_sf()
  
  #map
  map <- tm_shape(suitable_area_sf) +
    tm_fill(col = "area_km2", title = "Suitable Area (km²)", palette = palette) +
    tm_borders() +
    tm_layout(
      main.title = title,
      title.position = c("center", "top"),
      main.title.size = 1,
      legend.position = c("left", "bottom"), 
      legend.outside = TRUE,  
      legend.outside.position = "right",
      legend.outside.size = 0.5
    ) +
    tm_scale_bar(position = c("left", "bottom")) +
    tm_compass(position = c("right", "top"), size = 1)
  
  #table
  table <- suitable_area_sf %>%
    st_drop_geometry() %>% 
    select(rgn, area_km2) %>%
    arrange(desc(area_km2)) %>%
    kable(
      caption = "Suitable Habitat by Area (km²)",
      col.names = c("EEZ", "Suitable Area (km²)"),
      format = "html"
    ) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
  
  return(list(map = map, table = table))
}

```

### Run Oyster Function

```{r results = "asis"}
oyster_results <- suitable_locations(
  mean_sst = mean_sst, 
  depth_new = depth_new, 
  eez = eez, 
  sst_min = 11, 
  sst_max = 30, 
  depth_min = -70, 
  depth_max = -0, 
  title = "Suitable Oyster Aquaculture Areas"
)

print(oyster_results$map)
print(oyster_results$table)
```

### Run Abalone Fucntion

```{r results = "asis"}
abalone_results <- suitable_locations(
  mean_sst = mean_sst, 
  depth_new = depth_new, 
  eez = eez, 
  sst_min = 8, 
  sst_max = 18, 
  depth_min = -24, 
  depth_max = -0, 
  title = "Suitable Abalone Aquaculture Areas"
)

print(abalone_results$map)
print(abalone_results$table)
```

### Data Citations

| Suitable growing conditions | SeaLifeBase. 2024. Species Temperature and Depth Requirements for Marine Aquaculture. | https://www.sealifebase.ca/search.php |


| Sea surface temperatures | NOAA Coral Reef Watch. 2008-2012. 5km Daily Global Satellite Sea Surface Temperature Anomaly v3.1. | https://coralreefwatch.noaa.gov/product/5km/index_5km_ssta.php |


| Bathymetry | General Bathymetric Chart of the Oceans (GEBCO). 2024. Gridded Bathymetry Data. | https://www.gebco.net/data_and_products/gridded_bathymetry_data/#area |


| Exclusive economic zones | Marine Regions. 2024. Exclusive Economic Zones (EEZ) Boundaries, West Coast of the United States. | https://www.marineregions.org/eez.php |